{RandomRationalPoints} package for \emph{Macaulay2}

### 2020-10-03 FastLinAlg/Paper/RandomRationalPoints.tex
Changes throughout the paper.

\section{Helper functions}{\label{helper}}

All polynomial rings that we will consider will be over finite fields only.

We start by describing some functions that will come in handy for the main functions of this package.

\begin{itemize}
%	\item {\tt Codimension}: It takes input an integer. Various functions need to know the codimension/height of the scheme/ideal it is working with.  Setting this to be an integer will tell the function not to compute the codimension and to use this value instead.  The default value is {\tt null}, in which case the function will compute the codimension.  
%%%
%%Codimension is not a function, it is an option.
%%%
	\item {\tt randomCoordinateChange}: It takes a polynomial ring as an input and outputs the coordinate change map. 
	i.e. given a polynomial ring, this will produce a linear automorphism of the ring.

	In some applications, a full change of coordinates is not desired, as it might cause code to run slowly, and so a binomial change of coordinates might be preferred.  Alternately, we might only replace some coordinates by monomials.  
	Refinements of {\tt randomCoordinateChange} are controlled with the following options.

	\begin{itemize}
		\item {\tt MaxCoordinatesToReplace}: the user can specify that only a specified number of coordinates should be non-monomial.  Sometimes, generic coordinate changes where all coordinates are modified, can be very slow.  This is a way to mitigate for that.
		This option is passed to {\tt randomCoordinateChange} by other functions that call it. 		
		\item {\tt Replacement}: setting {\tt Replacement => Full} will mean that coordinates are changed to a general degree $1$ form.  If {\tt Replacement => Binomial}, the coordinates are only changed to binomials, which can be much faster for certain applications. If {\tt Homogeneous => false}, then there will be constant terms, and we view $mx + b$ as a binomial.
		\item 
	\end{itemize}

	  This function already checks whether the map is an isomorphism by computing the Jacobian.
\end{itemize}

@@

\vspace{1em}
\noindent where 
$I$ is an ideal 
in a polynomial ring, 
$R$ can denote a quotient of a polynomial ring and 
$n\in \mathbb{Z}$ is
an integer specifying how many dimensions to drop.

@@

\begin{example}
	\end{example}

Alternately, instead of {\tt I}, we may pass it a quotient ring.  It will then return the inclusion of the generic projection ring into the given ring, followed by the source of that inclusion.  It is essentially the same functionality as calling {\tt genericProjection(n, ideal R)} although the format of the output is slightly different. This method works by calling {\tt randomCoordinateChange} (see \Cref{helper}) before dropping some variables.  It passes the options {\tt Replacement}, {\tt MaxCoordinatesToReplace}, {\tt Homogeneous} to that function.

\begin{example}
	\end{example}

This function makes no attempt to verify that the projection is actually generic with respect to the ideal.

@@

\begin{example}
	content...
\end{example}
\end{enumerate}
\section{randomPoints}\label{randomPoints}
{\tt randomPoints} is a function to find rational points in a variety. The typical usages are as follows:

@@

\vspace{1em}
\noindent where $n$ is a positive integer denoting the number of desired points, and 
$I$ is an ideal inside a polynomial ring.

@@

\begin{itemize}
	\item {\tt BruteForce} simply tries random points and sees if they are on the variety.

@@

	\item {\tt Default} performs a sequence of different strategies, depending on the context.
\end{itemize}

@@

Various functions which produce points, or call functions which produce points, may naturally find scheme theoretic points that are not rational over the base field (for example, by intersecting with a random linear space).  Setting {\tt ExtendField => true} will tell the function that such points are valid.  Setting {\tt ExtendField => false} will tell the function ignore such points.  This sometimes can slow computation, and other times can speed it up.  In some cases, points over extended fields may also have better randomness properties for applications.

@@

This option is used by {\tt randomPoints} in some strategies to determine the maximum number of attempts to intersect with a linear space when looking for random rational points.  Other functions pass this option through to {\tt randomPoints}.

@@

As mentioned in the introduction, the two functions in this section will provide further tools for computing singular locus, in addition to those available in the package FastLinAlg.

@@

-- {\tt MinorPointAttempts => ZZ}: how many points to check the rank of the matrix at

@@

There is some similar functionality obtained via heuristics (as opposed to actually finding rational points) in the package ``FastLinAlg".

>>>>

\section{A helper function: randomCoordinateChange}{\label{helper}}

All polynomial rings that we will consider will be over finite fields.

We start {\tt randomCoordinateChange}, a helper function that also might be useful to others.

%	\item {\tt Codimension}: It takes input an integer. Various functions need to know the codimension/height of the scheme/ideal it is working with.  Setting this to be an integer will tell the function not to compute the codimension and to use this value instead.  The default value is {\tt null}, in which case the function will compute the codimension.  
%%%
%%Codimension is not a function, it is an option.
%%%
\subsection*{\tt randomCoordinateChange}: It takes a polynomial ring as an input and outputs the coordinate change map. 
	i.e. given a polynomial ring, this will produce a linear automorphism of the ring.  This function checks whether the map is an isomorphism by computing the Jacobian.

	In some applications, a full change of coordinates is not desired, as it might cause code to run very slowly.  A binomial change of coordinates might be preferred, or we might only replace some coordinates by monomials.  
	Refinements of {\tt randomCoordinateChange} are controlled with the following options.

	\begin{itemize}
		\item {\tt MaxCoordinatesToReplace}: the user can specify that only a specified number of coordinates should be non-monomial (assuming {\tt Homogeneous => true}).  
		This option is passed to {\tt randomCoordinateChange} by other functions that call it. 		
		\item {\tt Homogeneous}:  Setting {\tt Homogeneous => false} will cause degree zero terms to be added to modified coordinates (including monomial coordinates).
		\item {\tt Replacement}: setting {\tt Replacement => Full} will mean that coordinates are replaced by a general degree $1$ form.  If {\tt Replacement => Binomial}, the coordinates are only changed to binomials, which can be much faster for certain applications. If {\tt Homogeneous => false}, then there will be constant terms, and we view $mx + b$ as a binomial.
	\end{itemize}

	{{\small\color{blue}
	\begin{verbatim}
	i3 : R = ZZ/11[x,y,z];

	i4 : randomCoordinateChange(R)

           ZZ
o4 = map(R,--[x, y, z],{4x + 5y - 5z, 3x - 4y - 3z, 4x})                     
           11

	i5 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace=>1)

	o5 = | x -x-4y-5z y |

	i6 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace=>1, Homogeneous=>false)

	o6 = | x-3 z-5 -x+3y-4z+2 |

	i7 : matrix randomCoordinateChange(R, MaxCoordinatesToReplace=>1, Replacement=>Binomial)

	o7 = | y x+4z z |
	\end{verbatim}}
	}

@@

\vspace{1em}
\noindent where 
$I$ is an ideal 
in a polynomial ring, 
$R$ can denote a quotient of a polynomial ring and 
$n\in \mathbb{Z}$ is
an integer specifying how many dimensions to drop.  Note, this function makes no attempt to verify that the projection is actually generic with respect to the ideal.

@@

\begin{example}
	{{\small\color{blue}
	\begin{verbatim}
i1 : R=ZZ/5[x,y,z,w];

i2 : I = ideal(x,y^2,w^3+x^2);

i3 : genericProjection(2,I)

            ZZ                                         2          2
o6 = (map(R,--[z, w],{- x - 2y - z, - y - 2z}), ideal(z  - z*w - w ))                   
            5

o6 : Sequence
	\end{verbatim}
	}}
\end{example}

Alternately, instead of {\tt I}, we may pass it a quotient ring.  It will then return the inclusion of the generic projection ring into the given ring, followed by the source of that inclusion.  It is essentially the same functionality as calling {\tt genericProjection(n, ideal R)} although the format of the output is slightly different.

\begin{example}
	{{\small\color{blue}
	\begin{verbatim}
i1 : R=ZZ/5[x,y,z,w];

i2 : I = ideal(x,y^2,w^3+x^2);

i3 : genericProjection(2,R/I)

                            ZZ                          ZZ
                            --[z, w]                    --[z, w]                        
                R           5                           5
o3 = (map(----------------,--------,{2y - w, - y - z}), --------)                       
             2   3    2        4                            4
        (x, y , w  + x )      z                            z

o3 : Sequence
	\end{verbatim}
	}}
\end{example}

This method works by calling {\tt randomCoordinateChange} (see \Cref{helper}) before dropping some variables.  It passes the options {\tt Replacement}, {\tt MaxCoordinatesToReplace}, {\tt Homogeneous} to that function.

@@

\begin{example}
	{{\small\color{blue}
	\begin{verbatim}
i2 : R=ZZ/5[x,y,z,w];

i3 : I = ideal(x,y^2,w^3+x^2);

o3 : Ideal of R

i4 :genericProjection(2,I, Replacement=>Binomial)

            ZZ                                   4
o4 = (map(R,--[z, w],{x + 2z, - 2y + w}), ideal w )                                     
            5

o4 : Sequence
\end{verbatim}
}}
\end{example}

\end{enumerate}
\section{randomPoints}\label{randomPoints}
{\tt randomPoints} is a function to find rational or geometric points in a variety. The typical usages are as follows:

@@

\vspace{1em}
\noindent where $n$ is a positive integer denoting the number of desired points, and 
$I$ is an ideal inside a polynomial ring.  If {\tt n} is omitted, then it is assumed to be 1.

@@

\begin{itemize}
    \item {\tt Default} performs a sequence of the different strategies below, aimed at finding a point quickly.    
	\item {\tt BruteForce} simply tries random points and sees if they are on the variety.

@@

\end{itemize}

@@

Intersection with a general linear space will naturally find scheme theoretic points that are not rational over the base field.  
Setting {\tt ExtendField => true} will tell the function that such points are valid.  Setting {\tt ExtendField => false} will tell the function ignore such points.  This sometimes can slow computation, and other times can substantially speed it up when the variety has few rational points.  In some cases, points over extended fields may also have better randomness properties for applications.

@@

This option is used by {\tt randomPoints} in some strategies to determine the maximum number of attempts to intersect with a linear space when looking for random rational points.  %Other functions pass this option through to {\tt randomPoints}.

@@

As mentioned in the introduction, the two functions in this section will provide further tools for computing singular locus, in addition to those available in the package {\tt FastLinAlg}.

@@

-- {\tt MinorPointAttempts => ZZ}: the maximum number of points to check the rank of the matrix at

@@

There is some similar functionality obtained via heuristics (as opposed to actually finding rational points) in the package {\tt FastLinAlg}.
